Beating Simple Backtracking Search in Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1].}

### 2018-11-23 paper/paper.tex
Radical rewrite time

\author{Blair Archibald\orcidID{0000-0003-3699-6658}\and
Ciaran McCreesh\orcidID{0000-0002-6106-4871}\and
Patrick Prosser\orcidID{0000-0003-4460-6912} \and
James Trimble\orcidID{0000-0001-7282-8745}}

%\authorrunning{B. Archibald et al.}

\institute{University of Glasgow, Glasgow, Scotland\\
\email{ciaran.mccreesh@glasgow.ac.uk}}

@@

\begin{abstract}
    The current state of the art in subgraph isomorphism solving involves using degree as a
    value-ordering heuristic to direct backtracking search. Such a search makes a heavy commitment
    to the first branching choice, which is often incorrect. To mitigate this, we introduce and
    evaluate a new approach, which we call ``solution-biased search''. By combining a
    slightly-random value-ordering heuristic, rapid restarts, and nogood recording, we achieve two
    orders of magnitude faster performance on satisfiable instances, whilst not affecting
    performance on unsatisfiable instances. This search can also be parallelised in a very simple
    but effective way, giving us a further speedup of over ?? from thirty-six cores, across both
    satisfiable and unsatisfiable instances.  Finally, we show that solution-biased search is also
    suitable for optimisation problems, by using it to improve two maximum common induced subgraph
    algorithms.
\end{abstract}

@@

    \caption{Comparing depth-bounded discrepancy search to backtracking search, both with the degree
    heuristic.}\label{figure:scatter-dds}
\end{figure}

\Cref{figure:scatter-dds} shows the effects of adding Walsh's \cite{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to the solver (results with other discrepancy search variants
are similar).  Each point represents the solving time for one instance---to avoid noise for easier
instances, we measure only time spent during search, and exclude time spent in preprocessing and
initialisation.  Points below the $x-y$ diagonal are speedups, whilst points on the top and right
axes represent instances which timed out after one thousand seconds with one algorithm, but not the
other. For satisfiable instances, the different point styles show the different families, whilst all
unsatisfiable instances are shown as dark dots.  The points well below the diagonal line and along
the right-hand axis on the scatter plot show that DDS can sometimes be extremely beneficial on
satisfiable instances.  However, on both unsatisfiable and most satisfiable instances, the overheads
can be extremely large, and DDS is much worse in aggregate and is not a viable approach (even when
only considering satisfiable instances).  These large overheads are to be expected: discrepancy
searches are aimed primarily at getting better feasible solutions in optimisation problems which are
too large for a proof of optimality to be a realistic prospect, and they are not well-suited for
unsatisfiable decision problems. Despite this, the extremely large gains on some satisfiable
instances confirm our suspicions that we should find an alternative to heavy-commitment backtracking
search.

@@

?? Lead in

The left plot of \cref{figure:value-ordering-heuristics} shows the cumulative number of satisfiable
instances solved over time using different value-ordering heuristics.

\begin{description}
\item[Degree] We select vertices from highest degree to lowest degree
\cite{DBLP:conf/ijcai/McCreeshPT16}.

\item[Random] We select vertices uniformly at random.

\item[Anti] We select vertices from lowest degree to highest degree. This heuristic is used as
a sanity check: as expected, degree beats random, and random beats anti.

\item[Biased] We branch by selecting a vertex $v'$ from the chosen domain $D_v$ with
probability \[ p(v') = \frac{2^{\deg(v')}}{\sum_{w \in D_v}{2^{\deg(w)}}} \text{.} \]\end{description}

\noindent
The biased heuristic is new. It continues to prefer vertices of high degree, but will give the same
chance of being selected next to two vertices of equal degree.  It also introduces an element of
randomness, which will be needed to make restarts have an effect.  The performance of the degree and
biased heuristics are very similar: the biased heuristic has slightly worse performance below a 400s
timeout, and effectively equal performance above it. This shows that we can introduce an element of
randomness into the degree value-ordering heuristic without adversely affecting its performance
\emph{in aggregate} (and more generally, this scheme could be used for any value-ordering
heuristic which puts choices in some kind of rank with ties.)

@@

@@

?? Intuition

\subsection{Adding Restarts and Nogood Recording}

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in a special new \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve the solver by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis. Additionally, the solver's current
hand-crafted variable-ordering heuristics already beat adaptive heuristics like impact or
activity-based search.} Using restarts on value-ordering heuristics is uncommon (although
\citet{Razgon et al.\ }{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering heuristics from
restarts, \citet{Chu et al.\ }{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of parallel
search, and an early approach by \citet{Gomes et al.\ }{DBLP:conf/aaai/GomesSK98} does so in an
optimisation context).

@@

Preliminary experiments directed us to use the Luby scheme \cite{DBLP:journals/ipl/LubySZ93} to
determine when to restart.
%We count the number of backtracks
%(that is, when we reach the end of the main $\KwSty{foreach}$ loop on
%\siplineref{countrestarts})
%to decide when to restart, rather than counting recursive calls.
Following convention, we multiply each item in the Luby sequence by a constant---we used the SMAC
automatic parameter tuner \cite{DBLP:conf/lion/HutterHL11} to select the value 660.
% The remainder of the changes are to introduce nogoods.  The $B$ argument to $\FuncSty{search}$ tracks the branching choices
% made so far, in the form of a CNF clause with each literal being a guessed $(v \mapsto v')$
% assignment.
To avoid exploring portions of the search
space that we have already visited, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees---such a constraint is called a nogood. We generate
simple decision nogoods. That is, upon backtracking due to a decision to restart, we post a nogood
of the form $(v \mapsto v') \wedge (w \mapsto w') \wedge (x \mapsto x') \Rightarrow \bot$ for every
branch to the left of the current (incomplete) branch in the search tree, and when we first make a
decision to restart before backtracking, we post a similar nogood eliminating the entire subtree
explored.

Finally, we use the two watched literals technique \cite{DBLP:conf/dac/MoskewiczMZZM01} to
propagate stored nogoods.
%(\siplineref{2wl}).
This has two benefits: the propagation complexity does not particularly depend upon the number of
stored nogoods, and it does not require any work upon backtracking.  Other more sophisticated nogood
generation and propagation schemes exist \cite{DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17},
but these would not be helpful here since our solver does not maintain arc consistency or use a
propagation queue.

\subsection{The Effects of Restarts and Nogood Recording}

@@

    \caption{On the left, the number of satisfiable instances solved over time, with and without
    restarts and nogood recording. To the right, a comparison between the original algorithm and the
    new algorithm with all features enabled.}
    \label{figure:old-vs-new}
\end{figure}

In \cref{figure:old-vs-new} we show the effects of adding restarts and nogood recording to the
algorithm. With restarts and nogood recording, the random value-ordering heuristic comfortably beats
the degree strategy with simple backtracking search. In other words, although having a good
value-ordering heuristic is beneficial, introducing randomness into the search is better, if it is
done alongside a mechanism to avoid heavy commitment to any particular random choice.
\Cref{figure:old-vs-new} also shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored heuristic than it is to throw away the heuristic altogether.  Indeed, the
original algorithm can solve ?? satisfiable instances by ?? seconds, whilst the biased and
random restarting algorithms require only ?? seconds and ?? seconds respectively to solve the
same number.

In the more detailed view in the right-hand plot of \cref{figure:old-vs-new}, comparing the basic
algorithm with the degree heuristic to the degree-biased algorithm with restarts, all of the
unsatisfiable instances are very close to the $x-y$ diagonal, showing that their performance is
nearly unchanged. On the other hand, there are large numbers of satisfiable instances well below the
diagonal line, indicating large speedups.  Better yet, there are only a handful of satisfiable
instances that are more than a factor of ten times worse.  In other words, as well as improving
performance, we have made up most of the consistency we lost by introducing randomness.
This is a promising result: for the first time, we have introduced a viable alternative to
backtracking search in a setting where we both have a strong \emph{value}-ordering heuristic,
\emph{and} care about both satisfiable and unsatisfiable instances.

@@

>>>>

\author{Blair Archibald\inst{1}\orcidID{0000-0003-3699-6658}\and
Fraser Dunlop\inst{2}\orcidID{0000-0002-4485-4871}\and
Ruth Hoffmann\inst{2}\orcidID{0000-0002-1011-5894}\and
Ciaran McCreesh\inst{1}\orcidID{0000-0002-6106-4871}\and
Patrick Prosser\inst{1}\orcidID{0000-0003-4460-6912} \and
James Trimble\inst{1}\orcidID{0000-0001-7282-8745}}

\authorrunning{B. Archibald et al.}

\institute{University of Glasgow, Glasgow, Scotland \and University of St Andrews, St Andrews,
Scotland \\ \email{ciaran.mccreesh@glasgow.ac.uk}}

@@

\begin{abstract}
    The current state of the art in subgraph isomorphism solving involves using degree as a
    value-ordering heuristic to direct backtracking search. Such a search makes a heavy commitment
    to the first branching choice, which is often incorrect. To mitigate this, we introduce and
    evaluate a new approach, which we call ``solution-biased search''. By combining a
    slightly-random value-ordering heuristic, rapid restarts, and nogood recording, we achieve two
    orders of magnitude faster performance on satisfiable instances, whilst not affecting
    performance on unsatisfiable instances. This search can be parallelised in a very simple
    but effective way, giving us a further speedup of over ?? from thirty-six cores, across both
    satisfiable and unsatisfiable instances.  We also show that solution-biased search is also
    suitable for optimisation problems, by using it to improve two maximum common induced subgraph
    algorithms.
\end{abstract}

@@

    \caption{Comparing depth-bounded discrepancy search (DDS) to depth-first backtracking search
    (DFS), both using degree as the value-ordering heuristic.}\label{figure:scatter-dds}
\end{figure}

\Cref{figure:scatter-dds} shows the effects of adding Walsh's \cite{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to the solver (results with other discrepancy search variants
are similar). On the scatter plot, each point represents the solving time for one instance---to
avoid noise for easier instances, we measure only time spent during search, and exclude time spent
in preprocessing and initialisation.  Points below the $x-y$ diagonal are speedups, whilst points on
the top and right axes represent instances which timed out after one thousand seconds with one
algorithm, but not the other. For satisfiable instances, the different point styles show the
different families, whilst all unsatisfiable instances are shown as dark dots.  The points well
below the diagonal line and along the right-hand axis on the scatter plot show that DDS can
sometimes be extremely beneficial on satisfiable instances.  However, on both unsatisfiable and most
satisfiable instances, the overheads can be extremely large, and DDS is much worse in aggregate and
is not a viable approach (even when only considering satisfiable instances).  These large overheads
are to be expected: discrepancy searches are aimed primarily at getting better feasible solutions in
optimisation problems which are too large for a proof of optimality to be a realistic prospect, and
they are not well-suited for unsatisfiable decision problems. Despite this, the extremely large
gains on some satisfiable instances confirm our suspicions that we should find an alternative to
heavy-commitment backtracking search.

@@

@@

Traditionally, value-ordering heuristics are designed to drive search towards the most promising
region of the search space ?? cite, or the most constrained ?? cite, or the region with the highest solution
density ?? cite. In subgraph isomorphism, this is done by selecting vertices from highest degree to
lowest \cite{DBLP:journals/jair/McCreeshPST18}. The left-hand plot of
\cref{figure:value-ordering-heuristics} demonstrates that this is indeed a good choice: the
\textbf{Degree} heuristic's line shows much better performance on satisfiable instances than the
\textbf{Random} (branch randomly) or \textbf{Anti} (branch from lowest degree to highest) heuristic
lines. (On unsatisfiable instances, the value-ordering heuristic has no effect on performance,
because a complete search must be performed.)

But what happens if our value-ordering heuristic has to choose between mapping a pattern vertex to,
for example, three vertices of degree ten, two vertices of degree nine, or five of degree two? When
driving conventional backtracking search, the degree heuristic would pick one of the vertices of
degree ten, and we would commit all of our search effort to the exponentially large search tree
underneath it.  We will show that this is not a wise choice, and that instead, we should \emph{commit
equal search effort} to each of the three subproblems found by mapping to vertices of degree ten.
And similarly, should we be certain that vertex of degree ten is so much better than a vertex of
degree nine that we should commit no effort to degree nine vertices until all the degree ten
subproblems have been explored? Or might it be better to commit, say, twice as much effort to each
degree ten subproblems as to each degree nine subproblems, and only a very small amount of effort to
the degree two subproblems?  To test this hypothesis, we will now introduce a new alternative to
backtracking search, which we call \emph{solution-biased search}.  This search is made up of three
components: a new slightly-random value-ordering heuristic, rapid restarts, and nogood recording.

\subsection{Biased Value-Ordering}

We define a new \textbf{Biased} value-ordering heuristic, as follows. When branching, we select a
vertex $v'$ from the chosen domain $D_v$ with probability \[ p(v') = \frac{2^{\deg(v')}}{\sum_{w \in
D_v}{2^{\deg(w)}}} \text{.} \] This heuristic is now equally likely to pick between vertices of
equal degree, is twice as likely to pick a vertex of degree $d$ as one of degree $d - 1$, and is
over a thousand times more likely to pick a vertex of degree $d$ than degree $d - 10$.
\Cref{figure:value-ordering-heuristics} shows that the performance of the degree and biased
heuristics are very similar: the biased heuristic has slightly worse performance below a ??
timeout, and effectively equal performance above it. This shows that we can introduce an element of
randomness into the degree value-ordering heuristic without adversely affecting its performance
\emph{in aggregate}.

@@

\subsection{Restarting Search and Nogood Recording}

Having introduced a new value-ordering heuristic, we must now also move away from depth-first
backtracking search. We do this by using \emph{restarts} and \emph{nogood recording}. The general
idea is to perform a certain amount of search, and then if no solution has been found (and
unsatisfiability has not been proven), to abandon search and restart from the beginning. Such an
approach can only be beneficial if something changes after restarting---in a constraint programming
setting, this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in our new \emph{value}-ordering heuristic, and
continue to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may
be possible to further improve the solver by also introducing randomness or some form of learning
into its variable-ordering heuristic. However, simultaneously introducing a second change would
considerably complicate the empirical analysis. Additionally, the solver's current hand-crafted
variable-ordering heuristics already beat adaptive heuristics like impact or activity-based search.}
Using restarts on value-ordering heuristics is uncommon (although \citet{Razgon et al.\
}{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering heuristics from restarts, \citet{Chu
et al.\ }{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of parallel search, and an early
approach by \citet{Gomes et al.\ }{DBLP:conf/aaai/GomesSK98} does so in an optimisation context).

@@

Preliminary experiments directed us to use the Luby scheme \cite{DBLP:journals/ipl/LubySZ93} to
determine when to restart.
%We count the number of backtracks
%(that is, when we reach the end of the main $\KwSty{foreach}$ loop on
%\siplineref{countrestarts})
%to decide when to restart, rather than counting recursive calls.
Following convention, we multiply each item in the Luby sequence by a constant---we used the SMAC
automatic parameter tuner \cite{DBLP:conf/lion/HutterHL11} to select the value 660.
% The remainder of the changes are to introduce nogoods.  The $B$ argument to $\FuncSty{search}$ tracks the branching choices
% made so far, in the form of a CNF clause with each literal being a guessed $(v \mapsto v')$
% assignment.

To avoid exploring portions of the search
space that we have already visited, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees---such a constraint is called a nogood. We generate
simple decision nogoods. That is, upon backtracking due to a decision to restart, we post a nogood
of the form $(v \mapsto v') \wedge (w \mapsto w') \wedge (x \mapsto x') \Rightarrow \bot$ for every
branch to the left of the current (incomplete) branch in the search tree, and when we first make a
decision to restart before backtracking, we post a similar nogood eliminating the entire subtree
explored. We use the two watched literals technique \cite{DBLP:conf/dac/MoskewiczMZZM01} to
propagate stored nogoods.
%(\siplineref{2wl}).
This has two benefits: the propagation complexity does not particularly depend upon the number of
stored nogoods, and it does not require any work upon backtracking.  Other more sophisticated nogood
generation and propagation schemes exist \cite{DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17},
but these are not helpful in this setting (our solver does not maintain arc consistency or use a
propagation queue).

\subsection{Solution-Biased Search in Practice}

@@

    \caption{On the left, the number of satisfiable instances solved over time, comparing
    solution-biased search (SBS), random search with restarts (RSR), and the three value-ordering
    heuristics with conventional depth-first search. To the right, a comparison between the original
    algorithm and solution-biased search.}
    \label{figure:old-vs-new}
\end{figure}

In \cref{figure:old-vs-new} we show the effects of adding restarts and nogood recording to the
algorithm. With restarts and nogood recording (random search with restarts, RSR), the random
value-ordering heuristic comfortably beats the degree strategy with depth-first search. In other
words, although having a good value-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. However, the biased heuristic together
with restarts (solution-biased search, SBS) is better still---that is, if we are introducing
restarts, then it is better to add a small amount of randomness to a tailored heuristic than it is
to throw away the heuristic altogether.  Indeed, the original algorithm can solve ?? satisfiable
instances by ?? seconds, whilst the biased and random restarting algorithms require only ?? seconds
and ?? seconds respectively to solve the same number.

In the more detailed view in the right-hand plot of \cref{figure:old-vs-new}, comparing the original
algorithm to solution-biased search, all of the unsatisfiable instances are very close to the $x-y$
diagonal, showing that their performance is nearly unchanged. On the other hand, there are large
numbers of satisfiable instances well below the diagonal line, indicating large speedups.  Better
yet, there are only a handful of satisfiable instances that are more than a factor of ten times
worse.  In other words, as well as improving performance, we have made up most of the consistency we
lost by introducing randomness.  This is a promising result: for the first time, we have introduced
a viable alternative to backtracking search in a setting where we both have a strong
\emph{value}-ordering heuristic, \emph{and} care about both satisfiable and unsatisfiable instances.

@@

\subsection{Why ``Solution-Biased'', and why Exponential Bias?}

?? Although we have shown that it provides good results, we have yet to justify where the biased
formula comes from, or indeed why we call this approach ``solution-biased''. The crux of the
matter is that whilst we do not know where solutions actually lie in the search space, we \emph{do}
know that subproblems under a target vertex of degree $d$ tend to contain a small constant times
more search nodes than those under a target vertex of degree $d - 1$.
